---
title: "Reading Data"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Reading a `.csv` File

```{r}
library(tidyverse)
air_quality_data <- read_csv("https://www4.stat.ncsu.edu/online/datasets/AirQuality.csv")
air_quality_data
air_quality_data$`CO(GT)`[1:10]
```

# Reading in a Fixed Width Field (FWF) File

```{r}
help("read_fwf")
cigarettes_data <- read_fwf("https://www4.stat.ncsu.edu/online/datasets/cigarettes.txt", fwf_widths(c(17, 5, 9, 7, 4), c('brand', 'tar', 'nicotine', 'weight', 'co')), skip = 1)
cigarettes_data
```

# Manipulating Data with `dplyr`

```{r}
library(readr)
air_quality_data<- read_csv("https://www4.stat.ncsu.edu/online/datasets/AirQuality.csv")
#View(air_quality_data)
```

Manipulate this data and clean it up a bit

```{r}
air_quality_data |>
  select(-starts_with("...")) |>
  rename("co_gt" = `CO(GT)`,
         "pt_08_sq_co" = `PT08.S1(CO)`,
         "nmhc_gt" = `NMHC(GT)`,
         "c6h6_gt" = `C6H6(GT)`,
         "pt_08_s2_nmhc" = `PT08.S2(NMHC)`,
         "nox_gt" = `NOx(GT)`,
         "pt_08_s3_nox" = `PT08.S3(NOx)`,
         "no2_gt" = `NO2(GT)`,
         "pt_08_s4_no2" = `PT08.S4(NO2)`,
         "pt_08_s5_o3" = `PT08.S5(O3)`) |>
  filter(co_gt != -200) |>
  mutate(mean_co_gt = mean(co_gt, na.rm = TRUE)) |>
  View()
```

Now do the same creation of a new variable, but do it for multiple numeric variables using `across()`

```{r}
air_quality_data |>
  select(-starts_with("...")) |>
  rename("co_gt" = `CO(GT)`,
         "pt_08_sq_co" = `PT08.S1(CO)`,
         "nmhc_gt" = `NMHC(GT)`,
         "c6h6_gt" = `C6H6(GT)`,
         "pt_08_s2_nmhc" = `PT08.S2(NMHC)`,
         "nox_gt" = `NOx(GT)`,
         "pt_08_s3_nox" = `PT08.S3(NOx)`,
         "no2_gt" = `NO2(GT)`,
         "pt_08_s4_no2" = `PT08.S4(NO2)`,
         "pt_08_s5_o3" = `PT08.S5(O3)`) |>
  filter(co_gt != -200) |>
  mutate(across(where(is.numeric), mean, .names = "mean_{.col}")) |>
  View()
```

```{r}
air_quality_data |>
  select(-starts_with("...")) |>
  rename("co_gt" = `CO(GT)`,
         "pt_08_sq_co" = `PT08.S1(CO)`,
         "nmhc_gt" = `NMHC(GT)`,
         "c6h6_gt" = `C6H6(GT)`,
         "pt_08_s2_nmhc" = `PT08.S2(NMHC)`,
         "nox_gt" = `NOx(GT)`,
         "pt_08_s3_nox" = `PT08.S3(NOx)`,
         "no2_gt" = `NO2(GT)`,
         "pt_08_s4_no2" = `PT08.S4(NO2)`,
         "pt_08_s5_o3" = `PT08.S5(O3)`) |>
  filter(co_gt != -200) |>
  mutate(across(where(is.numeric), list(mean = mean, sd = sd), .names = "{.col}_{.fn}")) |>
  View()
```

Add in grouping functionality with `group_by()`

```{r}
air_quality_data |>
  select(-starts_with("...")) |>
  rename("co_gt" = `CO(GT)`,
         "pt_08_sq_co" = `PT08.S1(CO)`,
         "nmhc_gt" = `NMHC(GT)`,
         "c6h6_gt" = `C6H6(GT)`,
         "pt_08_s2_nmhc" = `PT08.S2(NMHC)`,
         "nox_gt" = `NOx(GT)`,
         "pt_08_s3_nox" = `PT08.S3(NOx)`,
         "no2_gt" = `NO2(GT)`,
         "pt_08_s4_no2" = `PT08.S4(NO2)`,
         "pt_08_s5_o3" = `PT08.S5(O3)`) |>
  filter(co_gt != -200) |>
  group_by(Date) |>
  mutate(across(where(is.numeric), list(mean = mean, sd = sd), .names = "{.col}_{.fn}")) |>
  View()
```


# Reading From a Database


```{r}
# need the bigquery
library(DBI)
library(tidyverse)

con <- dbConnect(
  bigrquery::bigquery(),
  project = "bigquery-public-data",
  dataset = "google_trends",
  billing = "st-558-498519"
)

dbListTables(con)

my_data <- tbl(con, "top_terms") |>
  select(refresh_date, term, rank, dma_id, dma_name) |>
  rename("Day" = "refresh_date", 
         "Top_Term" = "term") |>
  filter(rank == 1,
         Day > lubridate::as_date('2024-05-14'),
         dma_id == 500) |>
  collect()
```


# SQL Joins

```{r}
library(DBI)
library(tidyverse)
con <- dbConnect(RSQLite::SQLite(), "lahman.db")
dbListTables(con)
```

Combine the `Batting` table and the `Pitching` table on common variables

```{r}
#note this code differs slightly from what was in the video!
inner_join(tbl(con, "Batting") |> filter(yearID == 2000), 
           tbl(con, "Pitching") |> filter(yearID == 2000),
           by = c("playerID", "stint", "teamID", "lgID")) |>
  collect()
```


Can Write SQL code instead

```{r}
tbl(con, sql(
"SELECT p.playerID as pplayerID,
        p.stint as pstint,
        p.teamID as pteamID,
        p.lgID as plgID,
        p.G as pG,
        p.HR as pHR,
        p.BB as pBB,
        p.SO as pSO,
        p.HBP as pHBP,
        p.R as pR,
        p.SF as pSF,
        p.GIDP as pGIDP,
        p.IBB as pIBB,
        p.SH as pSH,
        p.W, p.L, p.GS, p.CG, p.SHO, p.SV, p.IPouts, p.ER, p.BAopp, 
        p.ERA, p.WP, p.BK, p.BFP, p.GF,
        b.*
FROM Pitching as p
INNER JOIN Batting as b on ((p.playerID = b.playerID) AND (pstint = b.stint) AND (pteamID = b.teamID) AND (plgID = b.lgID))
WHERE b.yearID = 2000 AND p.yearID = 2000"
))
```


Left join returns left table and matching right records

```{r}
left_join(tbl(con, "Batting") |> filter(yearID == 2000), 
           tbl(con, "Pitching") |> filter(yearID == 2000),
           by = c("playerID", "stint", "teamID", "lgID")) |>
  collect() |>
  select(playerID, ERA, everything())
```

Right join

```{r}
right_join(tbl(con, "Batting") |> filter(yearID == 2000),
           tbl(con, "Pitching") |> filter(yearID == 2000),
           by = c("playerID", "stint", "teamID", "lgID")) |>
  collect() |>
  select(playerID, ERA, everything())
```


Outer Join: All players are in the Batting table even if they have no at bats!

```{r}
full_join(tbl(con, "Batting") |> filter(yearID == 2000),
           tbl(con, "Pitching") |> filter(yearID == 2000),
           by = c("playerID", "stint", "teamID", "lgID")) |>
  collect()
```


