---
title: "Week11Notes"
format: html
---

# Creating an API with R

# APIs and Docker

APIs - a defined method for asking for information from a computer

- Useful for getting data

- Useful for allowing others to run your code and produce something

Docker - a 'container' to easily transport your code to others

- Include all relevant files (R, packages, etc.)

Ideas important for putting work into pipeline!

# Implementing a Model in Production

Need a way to make your model available to others

- Can write an API that accesses your model

- Not traditionally done in R!

- Often done in python or lower level language

- Can be done using the `plumber` package

# API Basics

`plumber` allows us to build a *REST API*

- Takes in an HTTP request from a client and outputs a response

  - Request and responses are specifically formatted

# `plumber` Package

Package creates API by you **decorating** your code with appropriate comments

- `#*`

Ex: Standard R function to paste a message together with comments prior

```{r}
#Send a message
#* @get /readme
function(){
  "This is our basic API"
}
#http://localhost:8000/readme
```

Let's try the template `plumber` file from RStudio!

- Endpoints are the logic defining what is returned from a request

- `@get`: a get request, specified by providing something like `/echo`

- Inputs are taken as strings by default

```{r}
#* Find natural log of a number
#* @param num Number to find ln of
#* @get /ln
function(num){
  log(as.numeric(num))
}
#http://localhost:PORT/ln?num=1
```

Another example of syntax

```{r}
#* Find multiple of two numbers
#* @param num1 1st number
#* @param num2 2nd number
#* @get /mult
function(num1, num2){
  as.numeric(num1)*as.numeric(num2)
}
#http://localhost:PORT/mult?num1=10&num2=20
```

Output is generally in JSON format

Can alter that:

- for HTML (\@serializer html)

- for PNG (\@serializer png), for JPEG (\@serializer jpeg)

- for htmlwidget (\@serializer htmlwidget)

Plot output as png

```{r}
#* Plot of iris data
#* @serializer png
#* @param type base or ggally
#* @param color TRUE or FALSE (only for ggally)
#* @get /plotiris
function(type = "base", color = FALSE){
  if(tolower(type) == "ggally"){
    if(color){
      a <- GGally::ggpairs(iris, aes(color = Species))
      print(a)
    } else {
      a <- GGally::ggpairs(iris)
      print(a)
    }
  } else {
    pairs(iris)
  }
}
#http://localhost:PORT/plotiris?type=base
```

- HTML widget as output

```{r}
#* Plotting widget
#* @serializer htmlwidget
#* @param lat latitude
#* @param lng longitude
#* @get /map
function(lng = 174.768, lat = -36.852){
  m <- leaflet::leaflet() |>
    addTiles() |>  # Add default OpenStreetMap map tiles
    addMarkers(as.numeric(lng), as.numeric(lat))
  m  # Print the map
}
#http://localhost:PORT/map?lng=174&lat=-36
```

# Basic API Process

- Create your R functions

- Load the `plumber` library

- Put them into a file to source with the `plumb` function (or use RStudio's button)

- Code way:

```{r}
library(plumber)
r <- plumb("API/myAPI.r")
r$run(port = 8000)
#is hosted at http://localhost:8000
```

# Recap

- Create our API via `plumber` package

- Decorate code with specific comments

- [Many other things](https://www.rplumber.io/) can be done we didn't cover!

# Docker Basics

# Docker Containers

Docker - a 'container' to easily transport your code/program to others

- Includes all relevant files (specific python install, specific packages, etc.)

- No chance of surprise bugs!

Run on a linux kernel

- `Dockerfile` specifies how to build a docker **image**

- Once built, you can run or deploy the image, creating a container

We'll just use pre-built images and run them locally on our machine!

# Steps to Using Docker

1.  Install docker desktop

2.  Download (pull) a base image

    - We'll use the `r-base` image to start

3.  Run the image to create a container

    - We'll do this locally via the command line (or terminal)!

    - Usually deploy containers on a cloud (like AWS).

    - Often done user a container management system like kubernetes

4.  Shut down the container

# 1. Installing Docker

Go to [Docker's website](https://docs.docker.com/get-docker/) to install `docker desktop`

- Once installed, open docker desktop (may take a few minutes to load)

- Open a **terminal**

  - Windows: Type `cmd` into your start menu

  - Mac: Go to Launchpad and type `terminal`

- Type `docker --version` and hit enter

- Type `docker run hello-world` and hit enter

# 2. Download a Docker Image

What's happening there?

Docker looks for an image called `hello-world` locally

- Didn't find it, downloads it from **docker hub**

- Runs a container (instance of that image) that prints its message to you

Download our image of interest via

- `docker pull r-base`

- Will take a minute to download! Once done

# 3/4. Creating our Container

- To run our image we can use `docker run _image_`

  - Open command prompt (with docker desktop running)

  - Submit `docker run -ti --rm r-base` to the console

  - `--rm` says remove the container when we close it

# Launch RStudio via Docker

- Let's go through an example where we load RStudio through docker!

Steps:

- Grab `rocker/rstudio` image

- Create container with

`docker run --rm -ti -e DISABLE_AUTH=true -p 127.0.0.1:8888:8888 rocker/rstudio`

- Use CTRL/CMD + C, `docker kill _container_name_`, or docker desktop to stop the container

<!-- -->

- May want [additional functionality via options](https://docs.docker.com/engine/reference/commandline/run/)! To access a local directory, we can mount the current working directory of our terminal:

  - (Windows via command prompt)\
    `docker run --rm -ti -e DISABLE_AUTH=true -p 127.0.0.1:8888:8888 %cd%:/home/rstudio/work rocker/rstudio`

  - (Mac via terminal)\
    `docker run --rm -ti -e DISABLE_AUTH=true -p 127.0.0.1:8888:8888 -v ${PWD}:/home/rstudio/work rocker/rstudio`

    - `-ti` allows us to run commands inside the container

    - `-p 127.0.0.1:8888:8888` exposes a port so we can access things via the browser

    - `-v path_to_folder:/home/rstudio/work` mounts a local folder for us to use in the container

    - `--rm` removes the container when we exit

    - `-e` sets an environment variable (so we don't have to log in)

## Useful Command Line Commands

- `cd` - change directory

- `cd ..` - move up a folder

- `ls` - list files in current directory

- `start .` or `open .` - open a folder view in your current directory

- `touch file_name.ext` - create a new file

- lots more!

## Useful Docker Commands to Know

- `docker pull _image_` - Download an image

- `docker image ls` - Check which images you have

- `docker run _image_` - Create an instance of the image (a container)

- `docker container ls` - Check which containers are currently running

- `docker kill _container_` - Stop a currently running container

# Recap

To avoid install issues when sharing software, we'll use everything through a docker container!

Common steps:

- Start docker desktop (wait for it to fully start!)

- `docker run ...`

- Open our web browser to the appropriate place

- Close command prompt/terminal to shut down container

# Building a Docker Image

# Docker

Suppose we want to build our own image!

- Need to create a `Dockerfile` telling Docker how to build our image

Let's create our API as a docker image

- First, create a directory for the relevant files/scripts

- Second, create a file called `Dockerfile` (no file extension) in that folder

Describes how to build the image

`FROM rstudio/plumber`

`RUN apt-get update -qq && apt-get install -y libssl-dev libcurl4-gnutls-dev libpng-dev libpng-dev pandoc`

`RUN R -e "install.packages(c('GGally', 'leaflet', 'plumber'))"`

`COPY myAPI.R myAPI.R`

`EXPOSE 8000`

`ENTRYPOINT ["R", "-e", \`\
`"pr <- plumber::plumb('myAPI.R'); pr$run(host='0.0.0.0', port=8000)"]`

# Build the Image

We now build the image

- Change your terminal's working directory to your folder with the dockerfile

Now build it: `docker build -t api .`

- `build` is telling docker we want to create an image

- `-t` indicates we are providing a name

- `api` is the name, `.` implies we are in the directory we want to build from

# Run the Container

Now we have the image, we use `docker run` to start it up!

`docker run --rm -p 8000:8000 api`

- `--rm` says remove the image when done

- `-p` gives docker access to our port

- `8000:8000` ports available to use (only 8000)

- `api` name of the image to run

# Other Things to Note

Should make sure things aren't still running once we stop using it

- `docker container ls` shows what containers are running

- `docker kill __name__` stops them

Can delete a container with `docker rm`

- `docker rm NAME`

When rebuilding, sometimes you don't want to use the cached builds

- `docker build -t NAME --no-cache`

Can export container with `docker export` (or image with `docker save`)!

- `docker export CONTAINER_NAME path.tar`

Can remove images with `docker rmi`

- `docker rmi -f ID`

# Recap

- We can build our own images!

- Create a `Dockerfile` and instruct it on what to do

- Need to Google some linux commands and do some troubleshooting!

# Dockerizing a Shiny App

# Docker

Recap

- You have a model, app, etc.

- You want to share it or make it available

Docker - a 'container' to easily transport your code to others

- Include all relevant files (R, packages, etc.)

- No chance of code not running, bugs, etc. (if you've done things right!)

# Dockerizing Shiny Apps

[Article 1](https://juanitorduz.github.io/dockerize-a-shinyapp/) and [Article 2](https://www.statworx.com/en/blog/how-to-dockerize-shinyapps/) on commands needed to Dockerize a shiny app

- If you can host your app somewhere, not really a need to dockerize...

# Dockerizing a Shiny App

- Create app in a folder with all files needed (images, etc)

  - All paths relative!

- Pull appropriate docker image

  - `docker pull rocker/shiny-verse`

- Create the `DOCKERFILE`

- Build the image

  - `docker build -t <NAME> <DOCKERFILE PATH>`

  - If in directory of `DOCKERFILE`, just put '.' for path

- Deploy the image

  - Ex: `docker run -p 3838:3838 <NAME>`

# Example Dockerization

Let's Dockerize a basic app

- An app to visualize bayes theorem
