---
title: "R Objects, Workflow, and Functions"
format: pdf
editor: visual
editor_options: 
  chunk_output_type: inline
---

# Vectors 

Create a vector

```{r}
set.seed(42)
my_unif <- runif(30)
is.vector(my_unif)
```




Subset that object

```{r}
my_unif[1:10]
my_unif[c(1:3, 15:17)]
```



Sort the vector

```{r}
sort(my_unif)
```


Create a vector with strings

```{r}
charVec <- c('aaf', 'Adf', 'fgh', 'hgf')
sort(charVec)
```



# Data Frames

- columns with the same type of data

```{r}
data(trees)
head(trees)
```

Subset a column

```{r}
trees$Girth
```


Get attributes from the data frame

```{r}
attributes(trees)
names(trees)
colnames(trees)[2:3]
```


# Lists 

Investigating data frame

```{r}
is.list(trees)
is.data.frame(trees)
```


Can subset as a list

```{r}
trees[1:2]
trees[[2]]
```


Look at linear model fit

```{r}
fit <- lm(Volume ~ Height + Girth, data = trees)
```



Look at structure but restrict info

```{r}
str(fit, max.level = 1)
```


Some helpful functions exist

```{r}
fit$coefficients
coef(fit)
fit$residuals
resid(fit)

fit$rank
#rank(fit) does not work
```


# if/then/else


Fizz buzz challenge

- take in a number
- if it's divisible by 3 return fizz
- if it's divisible by 5 return buzz
- if it's divisible by 15 return fizz buzz

```{r}
number <- 7

if((number %% 15) == 0) {
  print("fizz buzz")
} else if((number %% 5) == 0) {
  print("buzz")
} else if((number %% 3) == 0) {
  print("fizz")
} else {
  print("error")
}
  
  
#if(!(number %% 15))
```


# Loops

Wrap the fizz buzz code into a loop to check multiple values at once

```{r}
for(number in -1:15) {
  if((number %% 15) == 0) {
  print("fizz buzz")
  } else if((number %% 5) == 0) {
    print("buzz")
  } else if((number %% 3) == 0) {
    print("fizz")
  } else {
    print("error")
  }
}
```

# Writing Functions

Normal approximation to the binomial

```{r}
n <- 40
p <- .3

#probabilities from a binomial RV
dbinom(0:40, size = n, prob = p)

#plot 
plot(x = 0:n, y = dbinom(0:40, size = n, prob = p), type = 'h')

norm_x <- seq(from = 0, to = n, length = 1000)
lines(norm_x, dnorm(norm_x, mean = n*p, sd = sqrt(n*p*(1-p))))
```


Lets write a function to make this plot for any `n` and `p`

```{r}
plot_normApprox <- function(n,p) {
  
#plot 
  plot(x = 0:n, y = dbinom(0:n, size = n, prob = p), type = 'h')

  norm_x <- seq(from = 0, to = n, length = 1000)
  lines(norm_x, dnorm(norm_x, mean = n*p, sd = sqrt(n*p*(1-p))))
}
```


Test it out

```{r}
plot_normApprox(30, 0.5)
```


Add some default values

```{r}
plot_normApprox <- function(n = 30, p = 0.5) {
  
#plot 
  plot(x = 0:n, y = dbinom(0:n, size = n, prob = p), type = 'h')

  norm_x <- seq(from = 0, to = n, length = 1000)
  lines(norm_x, dnorm(norm_x, mean = n*p, sd = sqrt(n*p*(1-p))))
}
```


Test it

```{r}
plot_normApprox()
plot_normApprox(p = .1)
```








