---
title: "Week9Notes"
format: html
editor_options: 
  chunk_output_type: console
---

# Modeling Concepts

What is a (statistical) model?

- A mathematical representation of some phenomenon on which you've observed data

- Form of the model can vary greatly!

**Statistical learning** - Inference, prediction/classification, and pattern finding

- Supervised learning - a variable (or variables) represents an **output** or **response** of interest

  - May model response and

    - Make **inference** on the model parameters

    - **predict** a value or **classify** an observation

Our Goal: Understand what it means to be a good predictive model (not make inference)

```{r}
library(tidyverse)

bike_data <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/bikeDetails.csv")
bike_data <- bike_data |>
  mutate(log_selling_price = log(selling_price),
         log_km_driven = log(km_driven)) |>
  select(log_km_driven, log_selling_price, everything())
bike_data
```

```{r}
# SLR line
ggplot(bike_data, aes(x = log_km_driven, y = log_selling_price)) +
  geom_point() +
  geom_smooth(method = "lm")
```

```{r}
# add a squared x term
ggplot(bike_data, aes(x = log_km_driven, y = log_selling_price)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ x + I(x^2))
```

```{r}
# add a cubed x term
ggplot(bike_data, aes(x = log_km_driven, y = log_selling_price)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ x + I(x^2) + I(x^3))
```

```{r}
# using method = 'gam' (generalized additive model)
ggplot(bike_data, aes(x = log_km_driven, y = log_selling_price)) +
  geom_point() +
  geom_smooth()
```

```{r, eval = FALSE}
# tree based model
# never defined preds, won't run

ggplot(bike_data, aes(x = log_km_driven, y = log_selling_price)) +
  geom_point() +
  geom_line(data = preds, aes(x = log_km_driven, y = Tree_Predictions), color = "Blue",                     linewidth = 2)
```

## Training a Model

Once a class of models is chosen, we must define some criteria to fit (or train) the model

Simple Linear Regression (SLR) Model: $E(Y_i | x_i) = \beta_0 + \beta_1 x_i$

- Loss function - Criteria used to fit or train a model

  - For a given numeric response value, and prediction,

    - $y_i - \hat{y_i}$,

    - $(y_i - \hat{y_i})^2$,

    - $|y_i - \hat{y_i}|$

We try to optimize the loss over all the observations used for training

$$\sum_{i=1}^{n} (y_i - \hat{y_i})^2$$ $$\sum_{i=1}^{n} |y_i - \hat{y_i}|$$

### Fitting the SLR Model

- Often use squared error loss (least squares regression)

- Nice solutions for our estimates exist!

$$\hat{\beta_0} = \bar{y} + \bar{x} \hat{\beta_1}$$

$$\hat{\beta}_1 = \frac{\sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})}{\sum_{i=1}^{n} (x_i - \bar{x})^2}$$

```{r}
y <- bike_data$log_selling_price
x <- bike_data$log_km_driven
b1_hat <- sum((x-mean(x))*(y-mean(y)))/sum((x-mean(x))^2)
b0_hat <- mean(y)-mean(x)*b1_hat
c(round(b0_hat, 4), round(b1_hat, 4))
```

Now we can find a prediction! Denoted as $\hat{y}$

- Use `lm()` function to fit in R

- Utilizes `formula` notation: `y ~ x` -\> `response ~ model terms`

```{r}
slr_fit <- lm(log_selling_price ~ log_km_driven, data = bike_data)
slr_fit
```

## Inference Using SLR Model in R

If we assume iid errors that are Normally distributed with the same variance, we can conduct inference!

- Confidence intervals and hypothesis tests around the slope parameter

- Use summary() (generic function!) on our fitted model

```{r}
summary(slr_fit)
```

- Can use `anova()` to get Analysis of Variance information

```{r}
anova(slr_fit)
```

- Residual diagnostics can be found via `plot()` on the fitted model

```{r}
plot(slr_fit)
```

## Prediction Using SLR Model in R

- Can use the line for prediction with `predict()`!

  - Should supply fitted object and `newdata`

    - An optional data frame in which to look for variables with which to predict. If omitted, the fitted values are used

```{r, eval = FALSE}
predict
predict.lm
```

```{r}
predict(slr_fit, newdata = data.frame(log_km_driven = c(log(1000), log(10000), log(100000))))
```

```{r}
exp(predict(slr_fit, newdata = data.frame(log_km_driven = c(log(1000), log(10000), log(100000)))))
```

## Quantifying How Well the Model Predicts

We use a loss function to fit the model. We use a **metric** to evaluate the model!

- Often use the same loss function for fitting and as the metric

- For a given **numeric** response value, $y_i$, and prediction, $\hat{y_i}$,

$$(y_i - \hat{y_i})^2, |y_i - \hat{y_i}|$$

- Incorporate all points via

$$\frac{1}{n}\sum_{i=1}^{n}(y_i-\hat{y}_i)^2,\ \frac{1}{n}\sum_{i=1}^{n}\left|y_i-\hat{y}_i\right|$$

### Metric Function

- For a numeric response, we commonly use squared error loss as our metric to evaluate a prediction

$$L(y_i,\hat{y}_i)=(y_i-\hat{y}_i)^2$$

- Use **Root Mean Square Error** as a metric across all observations

$$\mathrm{RMSE}=\sqrt{\frac{1}{n}\sum_{i=1}^{n}L(y_i,\hat{y}_i)}=\sqrt{\frac{1}{n}\sum_{i=1}^{n}(y_i-\hat{y}_i)^2}$$

#### Commonly Used Metrics

For prediction (numeric response)

- Mean Squared Error (MSE) or Root Mean Squared Error (RMSE)

- Mean Absolute Error (MAE or MAD - deviation)

$$L(y_i,\hat{y}_i)=|y_i-\hat{y}_i|$$Huber Loss

- Doesn't penalize large mistakes as much as MSE

For classification (categorical response)

- Accuracy

- log-loss

- AUC

- F1 Score

### Evaluating our SLR Model

We could find our metric for our SLR model using the training data

- Called **training error**

```{r}
head(predict(slr_fit))
```

```{r}
mean((bike_data$log_selling_price-predict(slr_fit))^2)
```

```{r}
sqrt(mean((bike_data$log_selling_price-predict(slr_fit))^2))
```

Doesn't tell us how well we do on data we haven't seen!

## Training vs Test Sets

Ideally we want our model to predict well for observations it **has yet to see**!

- For *multiple* linear regression models, our training MSE will always decrease as we add more variables to the model...

- We'll need an independent test set to predict on (more on this shortly!)

## Big Picture Modeling

Supervised Learning methods try to relate predictors to a response variable through a model

- Lots of common models

  - Regression models

  - Tree based methods

  - Naive Bayes

  - k Nearest Neighbors

  - ...

- For a set of predictor values, each will produce some prediction we can call $\hat{y}$

- Evaluate model via a metric

- Will use an independent test set or cross-validation to more accurately judge our model

# Prediction & Training/Test Sets

In this course we focus on the **predictive modeling** paradigm. In the end, we’ll often fit different *families of models* (say some multiple linear regression models, a tree based model, and a random forest model) to a given data set. We’ll then judge those models using some model *metric* to determine which model is best at predicting!

Let’s break down this process into a few steps.

## Predictive Modeling Idea

First we’ll choose a form for a model. The types of models to consider often depends on the subject matter at hand.

Once we’ve chosen a model type, we **fit** the model using some algorithm. Usually, we can write this fitting process in terms of minimizing some **loss** function.

We then need to determine the quality of predictions made by the model. We use a model **metric** to do this. Quite often, the loss function and model metric are the same, but this isn’t always the case!

For numeric response, the most common model metric is mean squared error (MSE) or root mean squared error (RMSE). For a categorical reponse, the most commmon model metrics are accuracy and log-loss (discussed in detail later).

## **Training vs Test Sets**

Ideally we want our model to predict well for observations **it has yet to see**. We want to avoid *overfitting* to the data we train our model on.

The evaluation of predictions over the observations used to *fit or train the model* is called the **training (set) error**

- Let $y_i$ denote an observed value and $\hat{y}_i$ denote our prediction for that observation. If RMSE was our metric,

$$\mathrm{Training\ RMSE}=\sqrt{\frac{1}{\#\text{ of observations used to fit the model}}\sum_{\text{observations used to fit the model}}(y_i-\hat{y}_i)^2}$$

- If we only consider this error, we’ll have no idea how the model will fare on data it hasn’t seen!

One method to obtain a better idea about model performance is to *randomly* split the data into a **training set** and **test set**.

- On the training set we can fit (or train) our models

- We can then predict for the test set observations and judge effectiveness with our metric

- We want the training set to have about 80% - 90% of the total data

### **Example of Fitting and Evaluating Models**

```{r}
library(tidyverse)
bike_data <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/bikeDetails.csv")
bike_data <- bike_data |> 
  mutate(log_selling_price = log(selling_price), 
         log_km_driven = log(km_driven)) |>
  select(log_km_driven, log_selling_price, everything())
bike_data
```

Here our response variable is the `log_selling_price = ln(selling_price)`. We could consider the family of multiple linear regression (MLR) models with differing predictors (x variables).

- The basic MLR model with $p$ predictors models the average response variable given the predictors (that's what $E(Y \mid x_1, x_2, \ldots, x_p)$ represents, the average or expected $Y$, given (that's what the vertical bar means) the values of $x_1$, $x_2$, $\ldots$, $x_p$) as a linear function (linear in the parameter terms).

  $$E(Y \mid x_1, x_2, \ldots, x_p) = \beta_0 + \beta_1x_1 + \beta_2x_2 + \cdots + \beta_px_p$$

- $\beta_0$ is the intercept and each $\beta_j$ is a slope term associated with a predictor variable.

- As with SLR, we could model the error term with a Normal distribution if we were interested in making inference on the parameters.

These basic models find coefficients that minimize the sum of squared residuals (i.e. uses squared error loss to fit the model).

- MLR modeling fitting criterion (the ‘hats’ just imply the parameters are estimates rather than the ‘true’ underlying values):

$$\min_{\hat{\beta}\text{'s}} \sum_{i=1}^{n}\left(y_i-\left(\hat{\beta}_0+\hat{\beta}_1x_{1i}+\cdots+\hat{\beta}_px_{pi}\right)\right)^2$$

If you are familiar with maximum likelihood, the estimated $\beta$'s turn out to be equivalent to doing maximum likelihood estimation with the iid error Normal, constant variance, assumption!

- Consider three competing MLR models:

  - Model 1: log_selling_price = intercept + slope\*year + Error

  - Model 2: log_selling_price = intercept + slope\*log_km_driven + Error

  - Model 3: log_selling_price = intercept + slope\*log_km_driven + slope\*year + Error

We can split the data randomly into a training set and a testing set. There are a lot of ways to do this. We’ll use the `rsample::initial_split()` function.

- We commonly use an 80/20 or 70/30 training/test split. The proportion used in this split really depends on the amount of data you have and your subject matter expertise. More data in the test set means a better estimate of the model’s performance. However, less data in the training set means a more variable model (bigger changes in predictions from data set to data set).

- Let’s split our `bike_data` into a training and test set.

  - Use `initial_split()` to create an initial object

  - Use `training()` and `testing()` on that object to create the two data sets (note the number of observations in each set below!)

```{r}
library(tidymodels)
set.seed(10)
bike_split <- initial_split(bike_data, prop = 0.7)
bike_train <- training(bike_split)
bike_test <- testing(bike_split)
bike_train
```

```{r}
bike_test
```

We can fit or train these models on the training set. Recall we use `lm()` to easily fit an MLR model via formula notation. With formula notation we put our response variable on the left and our model for the predictors on the right. The model can include interactions, non-linear terms, etc. If we just want ‘main effects’ we separate predictors with + on the right hand side (we’ll cover this in more detail later!)

- Let’s fit our three models and save them as objects.

```{r}
reg1 <- lm(log_selling_price ~ year, data = bike_train)
coef(reg1)
```

```{r}
reg2 <- lm(log_selling_price ~ log_km_driven, data = bike_train)
coef(reg2)
```

```{r}
reg3 <- lm(log_selling_price ~ year + log_km_driven, data = bike_train)
coef(reg3)
```

ow we have the fitted models. Want to use them to predict the response

- Model 1: $\widehat{log\_selling\_price} = −186.1724+0.0978*year$

- Model 2: $\widehat{log\_selling\_price} = 14.6229−0.3899*log\_km\_driven$

- Model 3: $\widehat{log\_selling\_price} = -131.6674 + 0.0719*year − 0.2427*log\_km\_driven$

To get predictions from our model, we use the `predict()` function and specify the `newdata` we want to predict for as a data frame with column names matching our predictors in our models. If we don’t specify any `newdata`, it returns the predictions made on the training data.

- We can see the first few predictions on the training data from the first model via the code below

```{r}
bike_train |>
  mutate(mod1_preds = predict(reg1)) |>
  select(log_selling_price, year, mod1_preds)
```

Let’s use RMSE as our metric. Although not how we want to compare our models, we can obtain the **training** RMSE easily with `predict()`. Let’s code it up ourselves and also use `yardstick::rmse_vec()` to find it (this is the `tidymodels` way).

```{r}
#our own calculation for training RMSE
sqrt(mean((bike_train$log_selling_price - predict(reg1))^2))
```

- Now we can supply the actual responses and the model predictions to `yardstick::rmse_vec()`

```{r}
#using yardstick
yardstick::rmse_vec(bike_train$log_selling_price, predict(reg1))
```

```{r}
#second and third models
#using yardstick
rmse_vec(bike_train$log_selling_price, predict(reg2))
```

- These values represent a measure of quality of prediction by these models (as judged by our metric RMSE).

- This estimate of RMSE for the predictions is too **optimistic** compared to how the model would perform with new data!

- Really, what we want to compare is how the models do on data they weren’t trained on.

- We want to find this type of metric on the test set. That means we want to use the **truth** from the test set ($y_i$ for the test set) and compare that to predictions made for that test set observation ($\hat{y_i}$).

$$\mathrm{Test\ Set\ RMSE}=\sqrt{\frac{1}{n_{\text{test}}}\sum_{i=1}^{n_{\text{test}}}(y_i-\hat{y}_i)^2}$$ To do this in R we need to tell `predict()` about the `newdata` being the test set (`bike_test`). As this is a data frame with columns for our predictors, we can just pass the entire data frame for ease and `predict()` uses appropriate/needed columns to find our predictions.

```{r}
#look at a few observations and predictions
bike_test |>
  select(log_selling_price, year) |>
  mutate(model_1_preds = predict(reg1, newdata = bike_test))
```

- Now let’s find the **test set error** for each model

```{r}
#obtain the test set RMSE for each model
rmse_vec(bike_test$log_selling_price, predict(reg1, newdata = bike_test))
```

```{r}
rmse_vec(bike_test$log_selling_price, predict(reg2, newdata = bike_test))
```

```{r}
rmse_vec(bike_test$log_selling_price, predict(reg3, newdata = bike_test))
```

We see that our third model with both `year` and `log_km_driven` gives a better (by our metric) set of predictions!

When choosing a model, if the RMSE values were ‘close’, we’d want to consider the interpretability of the model (and perhaps the assumptions required by each model if inference was of interest!)

- Note: we can do this with `yardstick::rmse()` rather than `yardstick::rmse_vec()` if our predictions are in the data frame. `rmse()` takes in the `data` as the first argument, the `truth` column as the second, and the `estimate` (or predictions) as the third.

```{r}
#look at a few observations and predictions
bike_test |>
  select(log_selling_price, log_km_driven) |>
  mutate(model_1_preds = predict(reg1, newdata = bike_test)) |>
  rmse(truth = log_selling_price, estimate = model_1_preds)
```

## Recap

We generally need to go through a few steps when training and testing our model(s):

- Choose the form of the model

- Fit the model to the data using some algorithm

  - Usually this can be written as a problem where we minimize some loss function

- Evaluate the model using a metric

  - RMSE is a very common for a numeric response

- Ideally we want our model to predict well for observations **it has yet to see**!

# Cross Validation

We’ve discussed the idea of supervised learning. In this paradigm we have a **response** or **target** variable we are trying to predict.

- We posit some model for the response variable that is a function of our predictor variables.

- We fit that model to data (usually can be seen as minimizing some *loss* function).

- We judge how well our model does at predicting using some **metric.**

  - Usually we evaluate our metric on the test set after doing a training/test split.

Recall our example of fitting an MLR model from the last section of notes. There we first split the data into a training and test set. We fit the model on the training set. Then we looked at the RMSE on the test set to judge how well the models did.

$$\mathrm{RMSE}=\sqrt{\frac{1}{n_{\text{test}}}\sum_{i=1}^{n_{\text{test}}}(y_i-\hat{y}_i)^2}$$

where $y_i$ is the actual response from the test set and $\hat{y}_i$ is the prediction for that response value using the model.

**Note:** The terms metric and loss function are often used interchangeably. They are slightly different.

- **Loss function** is usually referred to as the criteria used to optimize or fit the model.

  - For an MLR model we usually fit by minimizing the sum of squared errors (MSE or RMSE equivalently) which is the same as using maximum likelihood under the iid Normal, constant variance, error model.

  - In this case, MSE or RMSE would be our loss function

- Our **metric** is the criteria for evaluating the model. When we have a numeric response, our most common criteria is RMSE! In this case the same as the common loss function for fitting the model.

We could fit our MLR model using a different loss function (say minimizing the absolute deviations). We could also evaluate our model using a different metric (say the average of the absolute deviations).

## **CV without a Training/Test Set Split**

Sometimes we don’t have a lot of data so we don’t want to split our data into a training and test set. In this setting, CV is often used by itself to select a final model.

Consider our data set about the selling price of motorcycles:

```{r}
bike_data <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/bikeDetails.csv")
bike_data <- bike_data |>
  mutate(log_selling_price = log(selling_price),
         log_km_driven = log(km_driven)) |>
  select(log_km_driven, year, log_selling_price, everything())
bike_data
```

- Let’s consider our three linear regression models for the `log_selling_price`

  - Model 1: log_selling_price = intercept + slope\*year + Error

  - Model 2: log_selling_price = intercept + slope\*log_km_driven + Error

  - Model 3: log_selling_price = intercept + slope\*log_km_driven + slope\*year + Error

- We can use CV error to choose between these models without doing a training/test split.

- We’ll see how to do this within the `tidymodels` framework shortly. For now, let’s do it by hand so we can make sure we understand the process CV uses!

### **Implementing 10 fold Cross-Validation Ourselves**

Let’s start with dividing the data into separate (distinct) folds.

1.  Split the data into 10 separate folds (subsets)

    - We can do this by randomly reordering our observations and taking the first ten percent to be the first fold, 2nd ten percent to be the second fold, and so on.

```{r}
nrow(bike_data)
```

```{r}
size_fold <- floor(nrow(bike_data)/10)
```

- Each fold, except one, will have 106 observations. There will be an extra observation in this case, we’ll lump that into our last fold.

- Let’s get our folds by using `sample()` to reorder the indices.

- Then we’ll use a for loop to cycle through the pieces and save the folds in a `list`

```{r}
set.seed(8)
random_indices <- sample(1:nrow(bike_data), size = nrow(bike_data), replace = FALSE)
#see the random reordering
head(random_indices)
```

```{r}
#create a list to save our folds in
folds <- list()
#now cycle through our random indices vector and take the appropriate observations to each fold
#you may want to set i to 1 and run the arguments to from and to in the seq() function
#change i to 2 and do the same thing to see how this works

for(i in 1:10){
  if (i < 10) {
    fold_index <- seq(from = (i-1)*size_fold + 1, to = i*size_fold, by = 1)
    folds[[i]] <- bike_data[random_indices[fold_index], ]
  } else { #for i = 10 we need to have 1 more observation 
    fold_index <- seq(from = (i-1)*size_fold + 1, to = length(random_indices), by = 1)
    folds[[i]] <- bike_data[random_indices[fold_index], ]
  }
}
folds[[1]]
folds[[2]]
```

- Let’s put this process into our own function for splitting our data up!

- This function will take in a data set and a number of folds (`num_folds`) and returns a list with the folds of the data.

```{r}
get_cv_splits <- function(data, num_folds){
  #get fold size
  size_fold <- floor(nrow(data)/num_folds)
  #get random indices to subset the data with
  random_indices <- sample(1:nrow(data), size = nrow(data), replace = FALSE)
  #create a list to save our folds in
  folds <- list()
  #now cycle through our random indices vector and take the appropriate observations to each fold
  for(i in 1:num_folds){
    if (i < num_folds) {
      fold_index <- seq(from = (i-1)*size_fold +1, to = i*size_fold, by = 1)
      folds[[i]] <- data[random_indices[fold_index], ]
    } else {
      fold_index <- seq(from = (i-1)*size_fold +1, to = length(random_indices), by = 1)
      folds[[i]] <- data[random_indices[fold_index], ]
    }
  }
  return(folds)
}
folds <- get_cv_splits(bike_data, 10)
```

2.  Now we can fit the data on nine of the folds and test on the tenth. Then switch which fold is left out and repeat the process. We’ll use MSE rather than RMSE as our metric for now.

```{r}
#set the test fold number
test_number <- 10
#pull out the testing fold
test <- folds[[test_number]]
test
```

```{r}
#remove the test fold from the list and use the rest as training data
train_list <- folds[-test_number]
#note that train is only length 9
length(train_list)
```

- Recall the `purrr:reduce()` function. This allows us to iteratively apply a function across an object. Here we’ll use `reduce()` with `rbind()` to combine the tibbles saved in the different folds.

```{r}
#combine the other folds into train set
train <- purrr::reduce(train_list, rbind)
#the other nine folds are now in train
train
```

- Now we can fit our model and check the metric on the test set

```{r}
fit <- lm(log_selling_price ~ log_km_driven, data = train)
#get test error (MSE - square RMSE)
(yardstick::rmse_vec(test[["log_selling_price"]], 
                     predict(fit, newdata = test)))^2
```

- Ok, let’s wrap that into a function for ease! Let `y` be a string for the response, `x` a character *vector* of predictors, and `test_number` the fold to test on.

```{r}
find_test_metric <- function(y, x, folds, test_number){
  #pull out the testing fold
  test <- folds[[test_number]]
  #remove the test fold from the list
  train_list <- folds[-test_number]
  #combine the other folds into train set
  train <- purrr::reduce(train_list, rbind)
  #fit our model. reformulate allows us to take strings and turn that into a formula
  fit <- lm(reformulate(termlabels = x, response = y), data = train)
  #get test error
  (yardstick::rmse_vec(test[[y]], predict(fit, newdata = test)))^2
}

#double check it works
find_test_metric(y = "log_selling_price", 
                 x = "log_km_driven", 
                 folds = folds, 
                 test_number = 10)
```

- We can use the `purrr::map()` function to apply this function to `1:10`. This will use each fold as the test set once!

```{r}
folds <- get_cv_splits(bike_data, 10)
purrr::map(1:10, find_test_metric, 
           y = "log_selling_price", 
           x = "log_km_driven", 
           folds = folds)
```

3.  Now we combine these into one value! We’ll average the MSE values across the folds and then take the square root of that to obtain the RMSE (averaging square roots makes less sense, which is why we found MSE first)!
    - We can again use `reduce()` to combine all of these values into one number! We average it by dividing by the number of folds and finally find the square root to make it the RMSE!

```{r}
sum_mse <- purrr::map(1:10, 
                      find_test_metric, 
                      y = "log_selling_price", 
                      x = "log_km_driven", 
                      folds = folds) |>
           reduce(.f = sum)
sqrt(sum_mse/10)
```

- Let’s put it into a function and make it generic for the number of folds we have.

```{r}
find_cv <- function(y, x, folds, num_folds){
  sum_mse <- purrr::map(1:num_folds, 
                        find_test_metric, 
                        y = y, 
                        x =x, 
                        folds = folds) |>
    reduce(.f = sum)
  return(sqrt(sum_mse/num_folds))
}
find_cv("log_selling_price", "log_km_driven", folds, num_folds = 10)
```

### **Compare Different Models Using Our CV Functions**

Now let’s compare our three separate models via 10 fold CV! We can use the same `folds` with different models. This will make our comparisons better (similar to using a block design if you are familiar with that).

```{r}
folds <- get_cv_splits(bike_data, 10)
```

1.  First let’s find our CV RMSE on our model with just `year` as a predictor:

```{r}
cv_reg_1 <- find_cv("log_selling_price", "year", folds, num_folds = 10)
```

2.  Now for our model with just `log_km_driven`:

```{r}
cv_reg_2 <- find_cv("log_selling_price", "log_km_driven", folds, num_folds = 10)
```

3.  And for the model with both of these (main) effects:

```{r}
cv_reg_3 <- find_cv("log_selling_price", c("year", "log_km_driven"), folds, num_folds = 10)
```

We can then use these CV scores to compare across models and choose the ‘best’ one!

```{r}
c("year" = cv_reg_1, "log_km_driven" = cv_reg_2, "both" = cv_reg_3)
```

We see that the model that has both predictors has the lowest cross-validation RMSE!

**We would now fit this ‘best’ model on the full data set!**

```{r}
best_fit <- lm(log_selling_price ~ year + log_km_driven, data = bike_data)
best_fit
```

## **CV Recap**

Nice! We’ve done a comparison of models on how well they predict on data they are not trained on without splitting our data into a train and test set! This is very useful when we don’t have a lot of data. Again, we would choose our best model (model 3 here) and refit the model on the full data set!

In some cases, we’ll want to use a train/test split and use CV on the training data only so we still have a separate test set for other comparisons. We’ll discuss this idea shortly! We’ll also see how to do this in the `tidymodels` framework rather than doing it by hand.

# Multiple Linear Regression

## Fitting MLR Models

- Create models with the same slope but intercepts differing by a categorical variable

```{r}
owner_fits <- lm(log_selling_price ~ owner + log_km_driven, data = bike_data)
coef(owner_fits)
```

- Create a data frame for plotting

```{r}
x_values <- seq(from = min(bike_data$log_km_driven),
                to = max(bike_data$log_km_driven),
                length = 2)
pred_df <- data.frame(log_km_driven = rep(x_values, 4),
                      owner = c(rep("1st owner", 2),
                                rep("2nd owner", 2),
                                rep("3rd owner", 2),
                                rep("4th owner", 2)))
pred_df <- pred_df |>
  mutate(predictions = predict(owner_fits, newdata = pred_df))
pred_df
```

- Plot our different intercept models

```{r}
ggplot(bike_data, aes(x = log_km_driven, y = log_selling_price, color = owner)) +
  geom_point() +
  geom_line(data = pred_df, aes(x = log_km_driven, y = predictions, color = owner))
```

- Create models with the different slopes and intercepts

```{r}
owner_fits_full <- lm(log_selling_price ~ owner*log_km_driven, data = bike_data)
coef(owner_fits_full)
```

- Plot our different intercept models

```{r}
ggplot(bike_data, aes(x = log_km_driven, y = log_selling_price, color = owner)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

Choosing an MLR Model:

- Given a bunch of predictors, tons of models you could fit! How to choose?

- Many variable selection methods exist...

- If you care mainly about prediction, just use cross-validation or training/test split!

  - Compare predictions using some metric!

  - We'll see how to use `tidymodels` to do this in a coherent way shortly!

# Modeling with `tidymodels`

## Convert Data

- We saw the use of `rsample::initial_split()`

  - If doing a non-learned transformation, do those first outside of tidymodels

```{r}
library(tidyverse)
library(tidymodels)
bike_data <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/bikeDetails.csv") |>
  mutate(log_selling_price = log(selling_price)) |>
  select(-selling_price)
#save creation of new variables for now!
bike_split <- initial_split(bike_data, prop = 0.7)
bike_train <- training(bike_split)
bike_test <- testing(bike_split)
```

- `initial_split()` allows for stratified sampling too!

## Data Preparation with `tidymodels`

`recipes` package within `tidymodels` allows for transformations

- Process keeps track of proper values to use for you!

- Start with `recipe()` call

  - Denote formula for response/predictors and data to use

  - `summary()` describes current setup (we don't want all of these as predictors)

```{r}
recipe(log_selling_price ~ ., data = bike_train) |>
  summary()
```

- `update_role()` allows you to declare types of variables (such as ID)

- This keeps the variable around even when not used in a model

```{r}
recipe(log_selling_price ~ ., data = bike_train) |>
  update_role(name, new_role = "ID") |>
  summary()
```

### Now Add Transformation Steps

Many `step_*` functions to consider

- `step_log()` to create our `log_km_driven` variable

- `step_rm()` to remove a variable

- `step_dummy()` to create dummy values for categorical variables

- `step_normalize()` to center and scale numeric predictors

```{r}
recipe(log_selling_price ~ ., data = bike_train) |>
 update_role(name, new_role = "ID") |>
 step_log(km_driven) |>
 step_rm(ex_showroom_price) |>#too many nas
 step_dummy(owner, seller_type) |>
 step_normalize(all_numeric(), -all_outcomes())
```

## `prep()` & `bake()` the Recipe

- If you have at least one preprocessing operation, `prep()` 'estimates the required parameters from a training set that can be later applied to other data sets'

- `bake()` applies the computations to data

```{r}
recipe(log_selling_price ~ ., data = bike_train) |>
  update_role(name, new_role = "ID") |>
  step_log(km_driven) |>
  step_rm(ex_showroom_price) |>
  step_dummy(owner, seller_type) |>
  step_normalize(all_numeric(), -all_outcomes()) |>
  prep(training = bike_train) |>
  bake(bike_train)
```

## Creating a Model with `tidymodels`

```{r}
linear_reg() %>%
 set_engine("lm") %>%
 translate()
```

Set up our model and recipes

```{r}
bike_rec <- recipe(log_selling_price ~ ., data = bike_train) |>
 update_role(name, new_role = "ID") |>
 step_log(km_driven) |>
 step_rm(ex_showroom_price) |>
 step_dummy(owner, seller_type) |>
 step_normalize(all_numeric(), -all_outcomes())

bike_mod <- linear_reg() %>%
 set_engine("lm")
```

## `workflow()`s with `tidymodels`

```{r}
bike_wfl <- workflow() |>
 add_recipe(bike_rec) |>
 add_model(bike_mod)
bike_wfl
```

## `fit()` That Model!

allows us to fit our model to a data set!

```{r}
bike_fit <- bike_wfl |>
 fit(bike_train)

bike_fit |>
 tidy()
```

## Find Test Set Metric(s)

```{r}
bike_wfl |>
 last_fit(bike_split) |>
 collect_metrics()
```

**The same transformations from the training set are used on the test set!**

## Fitting the Model with Cross-Validation

```{r}
bike_10_fold <- vfold_cv(bike_train, 10)
bike_CV_fits <- bike_wfl |>
 fit_resamples(bike_10_fold)

bike_CV_fits
```

Combine the metrics using `collect_metrics()`

```{r}
bike_CV_fits |>
 collect_metrics()
```

This is our CV error on the training set!

## Fit another Model with Cross-Validation for Comparison

```{r}
bike_int_rec <- recipe(log_selling_price ~ ., data = bike_train) |>
 update_role(name, new_role = "ID") |>
 step_log(km_driven) |>
 step_rm(ex_showroom_price) |>
 step_dummy(owner, seller_type) |>
 step_normalize(all_numeric(), -all_outcomes()) |>
 step_interact(terms = ~km_driven*year*starts_with("seller_type"))
```

Fit the model to the resamples and see our metric

```{r}
bike_int_CV_fits <- workflow() |>
 add_recipe(bike_int_rec) |>
 add_model(bike_mod) |>
 fit_resamples(bike_10_fold)
rbind(bike_CV_fits |> collect_metrics(),
 bike_int_CV_fits |> collect_metrics())
```

- Simpler model is better here

- Could now compare its performance on the test set to some other 'best' models

